The evolution of gene expression in the term placenta of viviparous mammals

### 2015-10-06 manuscript/manuscript.Rnw
add start of tree matrix plots

<<load.libraries,echo=FALSE,results="hide",warning=FALSE,message=FALSE,error=FALSE,cache=FALSE>>=
opts_chunk$set(dev="cairo_pdf",out.width="\\textwidth",out.height="0.7\\textheight",out.extra="keepaspectratio")
opts_chunk$set(cache=TRUE, autodep=TRUE)
options(device = function(file, width = 8, height = 7, ...) {
  cairo_pdf(tempfile(), width = width, height = height, ...)
})
options(scipen = -2, digits = 3)
library("data.table")
library("ggplot2")
library("reshape2")
library("xtable")
library("grid")
library("gridExtra")

@@

<<source.to.latex,echo=FALSE,results="hide",cache.extra=tools::md5sum(paste0("../common_r_code/",c("to_latex.R")))>>=
for (a in paste0("../common_r_code/",c("to_latex.R"))) {
    source(a)
}
@

<<load.data,echo=FALSE,results="hide",cache.extra=tools::md5sum(paste0("../data/",c("combined_fpkm")))>>=
for (a in paste0("../data/",c("combined_fpkm"))) {
    load(a)
}
star.logs[,value:=gsub("%$","",value)]
star.logs.wide <- 
    data.table(dcast(star.logs,species+file~field))
setnames(star.logs.wide,
         "Average input read length",
         "input.read.length")
setnames(star.logs.wide,
         "Number of input reads",
         "input.reads")
setnames(star.logs.wide,
         "Uniquely mapped reads number",
         "unique.map.reads")
@

@@

To verify expression levels in human and bonobo, we also generated
RNA-Seq data for the placenta transcriptome of the black-headed spider
monkey (Ateles fusciceps) using similar protocols discussed above.
Ateles is outside the clade, which includes hominids and can be used
to polarize expression values of highly expressed genes in Homo and
Pan. We sampled placental tissue from an Ateles placenta expelled from
the mother shortly after birth (see Supplemental Table X). We used
similar procedures discussed above to extract and sequence this
sample. [DEREK—I’m unsure how to proceed on describing this so as to
not step on Natalie’s toes]

@@

For transcriptomes sequenced with 75 bp, %
(\Sexpr{species(star.logs.wide[input.read.length > 70 & input.read.length < 80,unique(species)])})
%
,
the number of trimmed reads ranged from %
\Sexpr{star.logs.wide[input.read.length > 70 & input.read.length < 80,min(as.numeric(input.reads))]} 
%
to %
\Sexpr{star.logs.wide[input.read.length > 70 & input.read.length < 80,max(as.numeric(input.reads))]}
with at least 96\% of reads \DLA{This is wrong}
mapping to the respective reference genome (\fref{tab:ref_genome_aligned_species}).

For transcriptomes sequenced with 150 bp, %
% The mismatch here is because we're reporting trimmed reads here, not
% full length reads
(\Sexpr{species(star.logs.wide[input.read.length > 120 & input.read.length < 160,unique(species)])})
%
,
the number of trimmed reads ranged from %
\Sexpr{star.logs.wide[input.read.length > 120 & input.read.length < 160,min(as.numeric(input.reads))]} 
%
to %
\Sexpr{star.logs.wide[input.read.length  > 120 & input.read.length < 160,max(as.numeric(input.reads))]}
%
with at least 99\% of reads mapping to the
respective reference genome \DLA{This is also wrong}.
%
All reads were submitted to the NCBI Sequence Read Archive (SRA)
(Accession Number XXX-YYY).

For transcriptomes sequenced with 35 bp, %
% The mismatch here is because we're reporting trimmed reads here, not
% full length reads
(\Sexpr{species(star.logs.wide[input.read.length < 40 ,unique(species)])})
%
,
the number of trimmed reads ranged from %
\Sexpr{star.logs.wide[input.read.length < 40 ,min(as.numeric(input.reads))]} 
%
to %
\Sexpr{star.logs.wide[input.read.length  < 40  ,max(as.numeric(input.reads))]}
%
with at least 99\% of reads mapping to the
respective reference genome \DLA{This is also wrong}.
%
All reads were submitted to the NCBI Sequence Read Archive (SRA)
(Accession Number XXX-YYY).

@@

@@

@@

>>>>

<<load.libraries,echo=FALSE,results="hide",warning=FALSE,message=FALSE,error=FALSE,cache=FALSE>>=
opts_chunk$set(dev="cairo_pdf",out.width="\\textwidth",out.height="0.7\\textheight",out.extra="keepaspectratio")
opts_chunk$set(cache=TRUE, autodep=TRUE)
options(device = function(file, width = 8, height = 7, ...) {
  cairo_pdf(tempfile(), width = width, height = height, ...)
})
options(scipen = -2, digits = 3)
library("data.table")
library("ggplot2")
library("reshape2")
library("xtable")
library("grid")
library("gridExtra")
library("ggtree")

@@

<<source.to.latex,echo=FALSE,results="hide",cache.extra=tools::md5sum(paste0("../common_r_code/",c("to_latex.R","plot_tree_matrix.R")))>>=
for (a in paste0("../common_r_code/",c("to_latex.R","plot_tree_matrix.R"))) {
    source(a)
}
@

<<load.data,echo=FALSE,results="hide",cache.extra=tools::md5sum(paste0("../data/",c("combined_fpkm","gene_trees")))>>=
for (a in paste0("../data/",c("combined_fpkm","gene_trees"))) {
    load(a)
}
setkey(genes.to.tree,"gene_id")
star.logs[,value:=gsub("%$","",value)]
star.logs.wide <- 
    data.table(dcast(star.logs,species+file~field))
setnames(star.logs.wide,
         "Average input read length",
         "input.read.length")
setnames(star.logs.wide,
         "Number of input reads",
         "input.reads")
setnames(star.logs.wide,
         "Uniquely mapped reads number",
         "unique.map.reads")
@

@@

To verify expression levels in human and bonobo, we also generated
RNA-Seq data for the placenta transcriptome of the black-headed spider
monkey (\textit{Ateles fusciceps}) using similar protocols discussed above.
Ateles is outside the clade, which includes hominids and can be used
to polarize expression values of highly expressed genes in Homo and
Pan. We sampled placental tissue from an Ateles placenta expelled from
the mother shortly after birth (see Supplemental Table X). We used
similar procedures discussed above to extract and sequence this
sample. [DEREK—I’m unsure how to proceed on describing this so as to
not step on Natalie’s toes]

@@

For transcriptomes sequenced with 75 bp, %
(\Sexpr{species(star.logs.wide[input.read.length > 70 & input.read.length < 80,unique(species)])})
%
,
the number of trimmed reads ranged from %
\Sexpr{star.logs.wide[input.read.length > 70 & input.read.length < 80,min(as.numeric(input.reads))]} 
%
to %
\Sexpr{star.logs.wide[input.read.length > 70 & input.read.length < 80,max(as.numeric(input.reads))]}
with at least 96\% of reads \DLA{This is wrong}
mapping to the respective reference genome (\fref{tab:ref_genome_aligned_species}). 
%
For transcriptomes sequenced with 150 bp, %
% The mismatch here is because we're reporting trimmed reads here, not
% full length reads
(\Sexpr{species(star.logs.wide[input.read.length > 120 & input.read.length < 160,unique(species)])})
%
,
the number of trimmed reads ranged from %
\Sexpr{star.logs.wide[input.read.length > 120 & input.read.length < 160,min(as.numeric(input.reads))]} 
%
to %
\Sexpr{star.logs.wide[input.read.length  > 120 & input.read.length < 160,max(as.numeric(input.reads))]}
%
with at least 99\% of reads mapping to the
respective reference genome \DLA{This is also wrong}.
%
%
For transcriptomes sequenced with 35 bp, %
% The mismatch here is because we're reporting trimmed reads here, not
% full length reads
(\Sexpr{species(star.logs.wide[input.read.length < 40 ,unique(species)])})
%
,
the number of trimmed reads ranged from %
\Sexpr{star.logs.wide[input.read.length < 40 ,min(as.numeric(input.reads))]} 
%
to %
\Sexpr{star.logs.wide[input.read.length  < 40  ,max(as.numeric(input.reads))]}
%
with at least 99\% of reads mapping to the
respective reference genome \DLA{This is also wrong}.
%
All reads were submitted to the NCBI Sequence Read Archive (SRA)
(Accession Number XXX-YYY).

@@

\subsection{Spider Monkey Transcriptome}

\begin{wraptable}{L}{0.3\textwidth}
  \tiny
<<top.50.spider.monkey.expressed.genes,echo=FALSE,results="asis">>=
top.spider.monkey.genes <- top.genes.table(species.name="ateles fusciceps")
print(xtable(top.spider.monkey.genes),
      booktabs=TRUE,
      floating=FALSE
      )
setkey(top.spider.monkey.genes,"Gene name")
@ 
\caption{Expression in \ac{FPKM} of top 50 non-mitochondrial genes in
  \textit{Ateles fusciceps}}
\label{tab:top50_spider_monkey_placenta}
\end{wraptable}

@@

\begin{figure}
<<prolactin.evolution.tree,echo=FALSE,fig.width=8,fig.height=8>>=

tree.id <-
    genes.to.tree[combined.fpkm[gene_short_name=="PRL" & species=="homo sapiens",gene_id],tree]
gene.tree.table <- trees[[tree.id]]$table
gene.tree.expression <- combined.fpkm[gene_id %in% gene.tree.table[,gene_id]]
gene.tree <- read.tree(text=trees[[tree.id]]$tree.dat)
plot.tree.matrix(read.tree(text=trees[[tree.id]]$tree.dat),gene.tree.table,gene.tree.expression)
@ 
\caption{Expression of prolactin}
\label{fig:prolactin_evolution_tree}
\end{figure}

@@

\begin{figure}
<<galectins.evolution.tree,echo=FALSE,fig.width=8,fig.height=8>>=

tree.id <-
    genes.to.tree[combined.fpkm[gene_short_name=="LGALS1" & species=="homo sapiens",gene_id],tree]
gene.tree.table <- trees[[tree.id]]$table
gene.tree.expression <- combined.fpkm[gene_id %in% gene.tree.table[,gene_id]]
gene.tree <- read.tree(text=trees[[tree.id]]$tree.dat)
plot.tree.matrix(read.tree(text=trees[[tree.id]]$tree.dat),gene.tree.table,gene.tree.expression)
@ 
\caption{Expression of galectins}
\label{fig:galectins_evolution_tree}
\end{figure}
